Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-06-20 doc/invt.tex
invt.tex: small edits

In addition to the application to homogeneous updating schemes
(Appendix \ref{sec:homo_bandpass}),
bandpass updating schemes can also be applied to
a continuous variable, $z$,\footnote{
  In the transition to a continuous variable,
  we will map
  $\rho_i$ to $\rho(z) \delta z$,
  $\phi_{ki}$ to $\phi_k(z) \delta z$,
  $w_{ij}$ to $w(z, z') \delta z$, etc.}
with the eigenmodes derived from orthogonal polynomials, $R_k(z)$,
satisfying
\begin{align*}
  \int R_k(z) R_{k'}(z) \rho(z) \, d z = \delta_{k, k'}
  .
\end{align*}
%
With $\phi_k(z) = R_k(z) \, \rho(z)$,
we can readily satisfy Eqs. \eqref{eq:eig_orthonormal_cols}-\eqref{eq:ortho0},
and construct the updating matrix by
Eqs. \eqref{eq:w_from_phi} and \eqref{eq:lambda_bandpass} as
\begin{align*}
  w(z, z')
  =
  \sum_{k=1}^{K-1} \frac{ \phi_k(z) \, \phi_k(z') } { \rho(z) }
  = \sum_{k=1}^{K-1} R_k(z) \, R_k(z') \, \rho(z'),
\end{align*}
%
where we have dropped the $k=0$ mode
that corresponds to the uniform shift of the bias potential.
%
In this way, the bias potential is a superposition of the polynomials
$$
u(z, t) = \sum_{k=1}^{K-1} a_k(t) \, R_k(z),
$$
with the coefficients updated as, according to Eq. \eqref{eq:mbin_update},
$$
a_k(t+1) = a_k(t) + \alpha(t) \, R_k\bigl( z(t) \bigr).
$$

For example, by assuming the Legendre polynomials\cite{arfken}
on a flat distribution over $[z_c - \Delta z/2, z_c + \Delta z/2]$
and $K = 2$,
the bias potential contains only a linear mode for
$R_1(z) = 2 \sqrt{3} (z - z_c)/\Delta z$,
and we recover the algorithm by
Langfeld, Lucini and Rago\cite{langfeld2012}.
%
Applied to the energy space, $z = E$,
$a_1(t)$ represents the local or
statistical temperature\cite{kim2006, *kim2007},
and the inverse-time schedule
was recommended for its update\cite{pellegrini2014}.

Alternatively, by assuming the Hermite polynomials\cite{arfken}
on a Gaussian distribution centered at $z_c$ with width $\sigma_z$
and $K=3$,
we have a quadratic bias potential
as a linear combination of
$R_1(z) = (z-z_c)/\sigma_z$
and
$R_2(z) = (z - z_c)^2 /\left(\sqrt 2 \, \sigma_z^2\right)$.
%
This bias potential was commonly empolyed
in umbrella sampling (see e.g. Ref. \cite{zhu2012})
and the method of Gaussian ensemble\cite{hetherington1987,
*challa1988, *costeniuc2006, neuhaus2006, *neuhaus2007},
which is helpful in overcoming the barrier
in a first-order phase transition.
%
Initially, we can employ the WL strategy
to control the updating magnitude
with the histogram fluctuation redefined as
$$
\max\left\{
  \left|
    \frac{ \overline z - z_c } { \sigma_z }
  \right|
  ,
  \left|
    \frac{ \overline{ (z - z_c)^2 } - \sigma_z^2 }
    { \sqrt 2 \, \sigma_z^2 }
  \right|
  \right\}
  ,
$$
where the overline denotes the time average.

@@

We wish to show that with the inverse-time schedule,
the bias potential under the single-bin updating scheme
happens to coincide with a runtime average
of the independent corrections.
%as a natural consequence of the stabilization of a runtime average
%as the sample size increases\cite{
%  marsili2006, barducci2008}.
%
%We first note that a runtime average
%\begin{equation}
%  X(t) = \frac{x(1) + \cdots + x(t)}{t}
%  .
%  \label{eq:avx}
%\end{equation}
%can be written as a recurrence relation
%\begin{equation}
%  X(t) = X(t - 1) + \frac{x(t) - X(t-1)}{t}
%  .
%  \label{eq:avx_recur}
%\end{equation}
%%
%Here $x(t) - X(t-1)$ serves as an independent correction
%from data point at time $t$ to the previous average value,
%and this correction is absorbed into the average
%with a weight of $1/t$,
%which is the inverse sample size.

>>>>

In addition to the application to homogeneous updating schemes
(Appendix \ref{sec:homo_bandpass}),
bandpass updating schemes can also be applied to
a continuous variable, $z$,\footnote{
  In the transition to a continuous variable,
  we will map
  $\rho_i$ to $\rho(z) \delta z$,
  $\phi_{ki}$ to $\phi_k(z) \delta z$,
  $w_{ij}$ to $w(z, z') \delta z$, etc.}
with the eigenmodes derived from orthogonal polynomials, $R_k(z)$,
satisfying
\begin{align*}
  \int R_k(z) R_{k'}(z) \rho(z) \, d z = \delta_{k, k'}
  .
\end{align*}
%
With $\phi_k(z) = R_k(z) \, \rho(z)$,
we can readily satisfy Eqs. \eqref{eq:eig_orthonormal_cols}-\eqref{eq:ortho0},
and construct the updating matrix by
Eqs. \eqref{eq:w_from_phi} and \eqref{eq:lambda_bandpass} as
\begin{align*}
  w(z, z')
  =
  \sum_{k=1}^{K-1} \frac{ \phi_k(z) \, \phi_k(z') } { \rho(z) }
  = \sum_{k=1}^{K-1} R_k(z) \, R_k(z') \, \rho(z'),
\end{align*}
%
where we have dropped the $k=0$ mode
that corresponds to the uniform shift of the bias potential.
%
This results in a polynomial bias potential
that can be written as a superposition
$$
u(z, t) = \sum_{k=1}^{K-1} c_k(t) \, R_k(z),
$$
with the coefficients updated as, according to Eq. \eqref{eq:mbin_update},
$$
c_k(t+1) = c_k(t) + \alpha(t) \, R_k\bigl( z(t) \bigr).
$$

For example,
to sample a flat distribution, $\rho(z)$,
over $[z_c - \Delta z/2, z_c + \Delta z/2]$,
we can use the Legendre polynomials\cite{arfken},
and with $K = 2$,
the bias potential contains only a linear mode for
$R_1(z) = 2 \sqrt{3} (z - z_c)/\Delta z$,
which recovers the algorithm by
Langfeld, Lucini and Rago\cite{langfeld2012}.
%
Applied to the energy space, $z = E$,
$c_1(t)$ represents the scaled local or
statistical temperature\cite{kim2006, *kim2007},
and the inverse-time schedule
was recommended for its update\cite{pellegrini2014}.

Alternatively,
to sample a Gaussian distribution
centered at $z_c$ with width $\sigma_z$
we can use the Hermite polynomials\cite{arfken},
and with $K=3$,
we have a quadratic bias potential
from a linear combination of
$R_1(z) = (z - z_c)/\sigma_z$
and
$R_2(z) = (z - z_c)^2 /\left(\sqrt 2 \, \sigma_z^2\right)$.
%
This bias potential was commonly empolyed
in umbrella sampling (see e.g. Ref. \cite{zhu2012})
and the method of Gaussian ensemble\cite{hetherington1987,
*challa1988, *costeniuc2006, neuhaus2006, *neuhaus2007},
which is helpful in overcoming barriers
in first-order phase transitions.
%
While the asymptotic updating magntiude
would be given by the inverse-time formula,
we can employ the WL strategy early in simulation
to control the updating magnitude in stages
with the histogram fluctuation redefined as
$$
\max\left\{
  \left|
    \frac{ \overline z - z_c } { \sigma_z }
  \right|
  ,
  \left|
    \frac{ \overline{ (z - z_c)^2 } - \sigma_z^2 }
    { \sqrt 2 \, \sigma_z^2 }
  \right|
  \right\}
  ,
$$
where the overline denotes the time average.

With only two nontrivial modes,
the above strategy usually only works
for a Gaussian with a narrow width.
%
We can, however, use it as a building block
to construct an expanded ensemble with
multiple Gaussians centered at different places.
%
Such a construction can be realized via
either the replica exchange
(parallel tempering)\cite{swendsen1986,
*geyer1991, *hukushima1996, *hansmann1997}
or the simulated tempering\cite{marinari1992,
  *lyubartsev1992, li2007,
  park2007, *nguyen2013, *zhang2015st}
framework.
While the mutual exchange probabiliy
can be readily derived in the former
framework\cite{neuhaus2006, *neuhaus2007},
the transition probability in the latter case is
requires estimation of the partition function
of the Gaussian ensemble
$Z \equiv \int e^{-\beta \, E(\mathbf x)-u(z(\mathbf x),t)} \, d\mathbf x$,
where $\mathbf x$ denotes the molecular coordinates,
$E(\mathbf x)$ is the unbiased energy function,
$\beta$ is the inverse temperature,
and the collective variable, $z$,
has also been written as a function of $\mathbf x$.
%
The transition probabiliy from
the $i$th Gaussian to the $j$th is given by
is given by
\begin{align*}
A(i\to j) =
\min\left\{1,
  \frac{Z^{(i)}}{Z^{j}}
  \exp\left[u^{(i)}\bigl( z(t) \bigr)
   - u^{(j)}\bigl( z(t) \bigr) \right]
  \right\}
  ,
\end{align*}
where $z(t)$ is the current value of $z$.
%
For a neighboring pair, $j = i\pm 1$,
we can estimate the ratio $Z^{(i)}/Z^{(j)}$
from the following approximate formula
%
\begin{align*}
  \frac{Z_i}{Z_j}
  =
  \frac{\sigma_i}{\sigma_j}
  \exp\left[
    u^{(j)}\bigl( z^{(i,j)} \bigr)
   -u^{(i)}\bigl( z^{(i,j)} \bigr)
  \right]
  ,
\end{align*}
%
where
$z^{(i,j)} \equiv
\bigl( z_c^{(i)}/\sigma_z^{(i)} + z_c^{(j)}/\sigma_z^{(j)} \bigr) /
\bigl(         1/\sigma_z^{(i)} +         1/\sigma_z^{(j)} \bigr)$.

@@

We wish to show the optimality of the inverse-time schedule
for the single-bin updating scheme
from the observation that
it leads to a bias potential
that coincides with a runtime average
of the independent corrections.
%as a natural consequence of the stabilization of a runtime average
%as the sample size increases\cite{
%  marsili2006, barducci2008}.
%
%We first note that a runtime average
%\begin{equation}
%  X(t) = \frac{x(1) + \cdots + x(t)}{t}
%  .
%  \label{eq:avx}
%\end{equation}
%can be written as a recurrence relation
%\begin{equation}
%  X(t) = X(t - 1) + \frac{x(t) - X(t-1)}{t}
%  .
%  \label{eq:avx_recur}
%\end{equation}
%%
%Here $x(t) - X(t-1)$ serves as an independent correction
%from data point at time $t$ to the previous average value,
%and this correction is absorbed into the average
%with a weight of $1/t$,
%which is the inverse sample size.
